---
---
# Intro to Spatial Data

This workshop borrows largely from Noli Brazil's class [CRD 230: Spatial Methods in Community Research](https://crd230.github.io)

This workshop will cover some useful packages to wrangle and visualize spatial data. Specifically, we will learn how to pull in data from a few different sources, set a Coordinate Reference System, and create a map. 

The example here mostly works with vector data, which is characterized by real world objects like a tree, building, forest, etc. The two main packages that deal with vector data in R:

-   **sp:** the original spatial package in R (first released in 2005) that has a complex data structure

-   **sf:** much newer, so not as well established as **sp**, but it is easier to use than sp and is tidy friendly

SF stands for "simple feature". It sees any real-word object (or vector) as a feature and a feature can have both a geometry of where the feature is located and attributes associated with it. For example, a county is a feature with a polygon shape and can have attributes like population, number of hospitals, etc. associated with it. Another example could be a UC Davis, which would have a point geometry.

Although, this workshop mostly covers vector data, it will also touch on another type of spatial data called raster data. Raster data is made up of pixels/cells that form a grid filled with information, that is typically continuous in nature (e.g., elevation temperature, land use). 
# Spatial Data Workflow

Here is a sample spatial data workflow: 
1. Bring in spatial data of interest into R
2. Wrangle data into desired format (e.g., intersect/within/clip, merge multiple spatial dataframes, areal interpolation, etc.)
3. Set Coordinate Reference System (and potentially reproject or specify measurement units)
4. Map data

# Mapping trees in Davis Example 
I am curious about the spatial distribution of fruit trees in Davis, and how this distribution pattern relates to population size and median housing value in Davis. In order to examine this, we will need to bring in US Census data for the City of Davis, as well as point data from fallingfruit.org

## Bring in data

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, results = "hide", error = FALSE, warning = FALSE)

library(tidyverse)
library(rmapshaper) # spatial data wrangling
library(tigris) #brings in places data
library(tidycensus) #brings in census data
library(sf) #spatial package
library(tmap) #mapping package
library(bookdown)
library(leaflet)#mapping package
```

Bring in all datasets of interest. In our case, we are going to use three sources of data: 1) City of Davis boundary, 2) census tract data for population and median housing value, and 3) fallenfruit.org fruit tree locations.

```{r census_tracts}
# You will need to sign up for a Census API Key if you are interested in pulling in US Census data 
## Request an API key here: https://api.census.gov/data/key_signup.html
#census_api_key("3f1a61d7b9d2f870de53940d461dda896938261b", install = TRUE) #now stored in my R environment

# The tigris packages provides a way to directly download incorporated city footprints
## Pull in in places data for California
pl <- places(state = "CA", year = 2020, cb = FALSE)

## Now let's just isolate the boundary for the City of Davis
davis <- pl %>%
  filter(NAME == "Davis")

## Take a look at the dataframe
glimpse(davis)

# Let's also pull in some census tract data from the City of Davis
## First, let's look at what variables are available 
v20 <- load_variables(2020, "acs5", cache=TRUE) # load variable options

## Now that we have chosen our variables of interest, let's pull in that data
ca.tracts <- get_acs(
      geography = "tract",
      year = 2020, # final year
      variables = c(totp = "B01003_001", #median income
                    medhouse = "B25077_001"), #Median housing value for owner-occupied housing units
      state = "CA",
      output= "wide",
      survey = "acs5", #this loads the data from the last 5 years of acs records
      geometry = TRUE,
      cb = FALSE
      )

# Now bring in fallingfruit.org
## I went onto the website and directly downloaded the csv file
fruit <- read.csv("data/data.csv")
```

## Data wrangling 
Oftentimes, you may want to subset spatial data relative to other spatial data. Some common spatial data wrangling tasks: 

Intersect: keeps all polygons that intersect with the specified spatial boundary

Within: keeps all polygons that are wholy within the specified spatial boundary

Clipping: clips polygons based on specific spatial extents

Areal interpolation: "allocation of data from one set of zones to a second overlapping set of zones that may or may not perfectly align spatially" 

For this example, I am only interested in the distribuition of fruit trees within Davis. The ms_clip function from rmapshaper is useful for clipping data to the spatial extents of Davis.

```{r clip}
## Now I am only interested in Davis census tracts, so I will use the city of davis boundary to clip the tracts
## This function comes from the rmapshaper package
davis.tracts <- ms_clip(target = ca.tracts, clip = davis, remove_slivers = TRUE) 
```
## Coordinate Reference System

Now that we have all of our data, we need to make sure that each dataset has the same Coordinate Reference System (CRS). The CRS has two parts:

1.  Geographic Coordinate System (GCS): three dimensional spherical surface. The GCS is made up of the ellipse (how the earth's roundness is calculated) and the datum (coordinate system).
2.  Projected Coordinate System (PCS), or "projection": Flattens the GCS into a two-dimensional

Both GCS and PCS need to be specified when working with spatial data! In order to do this, you will first need to find out what CRS your spatial dataset was created in initially.

```{r crs}
# Check CRS of all three datasets
class(davis)
st_crs(davis) #NAD83
st_crs(davis)$proj4string #"+proj=longlat +datum=NAD83 +no_defs" 
st_crs(davis)$units #NULL

st_crs(davis.tracts) #NAD83
st_crs(davis.tracts)$proj4string #"+proj=longlat +datum=NAD83 +no_defs" 
st_crs(davis.tracts)$units #NULL

st_crs(fruit) #NA
st_crs(fruit)$proj4string #NA

# So we can see that both the davis and ca.tracts dataframes are in NAD 83, so let's reproject the fruit dataframe to also be in this projection

# Reprojection
## First, establish the CRS for the fruit dataset based on how it was created intially
## Since fruit is point data, the projected coordinate system is already set because latitude and longitude are the X-Y coordinates but we need to tell R this
fruit.sf <- fruit %>%
  st_as_sf(coords = c("lng", "lat"), 
           crs = "+proj=longlat +datum=WGS84 +ellps=WGS84")
st_crs(fruit.sf)# +proj=longlat +datum=WGS84 +ellps=WGS84 

## Now let's reproject so it is on the same coordinate system as the other dataframes
### By transform to proj = utm, now the CRS can handle distance measures
### UTM: Universal Transverse Mercator works in meters
fruit.utm <- fruit.sf %>%
  st_transform(crs = "+proj=utm +zone=10 +datum=NAD83 +ellps=GRS80")

## Reproject davis and ca.tracts to also be in UTM
davis.utm <- davis %>%
  st_transform(crs = "+proj=utm +zone=10 +datum=NAD83 +ellps=GRS80")

davis.tracts.utm <- davis.tracts %>%
  st_transform(crs = "+proj=utm +zone=10 +datum=NAD83 +ellps=GRS80")

# Great, nows lets check to see if all dataframes are on the same CRS
st_crs(fruit.utm) == st_crs(davis.utm) #TRUE

st_crs(davis.utm) == st_crs(davis.tracts.utm) #TRUE


```

## Mapping the data
Finally, we are at the point where we can map the data! Let's first see what mapping it with tmap looks like.

```{r tmap}
# Map the point data over the census tract data
## tmap 
tm_shape(davis.tracts.utm) +
  tm_polygons(col = "medhouseE", style = "quantile", palette = "Blues",
              title = "Median Housing Value ($)") +
  tm_shape(fruit.utm) +
  tm_dots(col = "green") +
  #tm_text("types") +
    tm_scale_bar(breaks = c(0, 1, 2), text.size = 1, position = c("left", "bottom")) +
      tm_compass(type = "4star", position = c("right", "bottom"))  +
  tm_layout(main.title = "Fruit Trees in Davis", 
            main.title.size = 1.25, main.title.position="center",
            legend.outside = TRUE, legend.outside.position = "right",
            frame = FALSE)

```

Now let's try mapping with leaflet. 

```{r leaflet}
## need to reproject davis.tracts data to be +proj=longlat
davis.tracts.sf <- davis.tracts %>%
  st_transform(crs = "+proj=longlat +datum=WGS84 +ellps=WGS84")

leaflet() %>%
  addTiles() %>%
  addMarkers(data = fruit.sf, popup = ~as.character(types), label = ~as.character(types)) %>%
  addPolygons(data = davis.tracts.sf, 
              color = ~colorQuantile("Blues", totpE, n = 5)(totpE),
              weight = 1,
              smoothFactor = 0.5,
              opacity = 1.0,
              fillOpacity = 0.5,
              highlightOptions = highlightOptions(color = "white",
                                                  weight = 2,
                                                  bringToFront = TRUE))
```
